### 2016-08-02 forgerard.tex
writing the note

\begin{document}
The model for the intrinsic absolute magnitudes $U$, $B$, $V$, $R$, and $I$ is that they are drawn from a Normal distribution
\begin{equation*}
\left(
\begin{matrix}
U\\B\\V\\R\\I
\end{matrix}
\right) \sim \mathcal{N}
\left(
\Delta +
\left(
\begin{matrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} \\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si}  \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} \\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} \\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}
\end{matrix}
\right)
,C_{c}
\right).
\end{equation*}
The mean of the normal distribution linearly relates absolute magnitudes with
individual spectral parameters  $EW_{Ca}$, and $EW_{Si}$ via the  global coefficients
 $c_i$, $\alpha_i$, $\beta_i$.  Residual variations in absolute magnitude is described
 by  $C_c$ the intrinsic dispersion of the Normal distribution.  To constrain the problem we
 set $\langle \Delta \rangle=0$.

The observed magnitudes  $U_o$, $B_o$, $V_o$, $R_o$, and $I_o$  and spectral features  $EW_{Ca,o}$, and $EW_{Si,o}$ are modeled
as being drawn from a Normal distribution. 
\begin{equation*}
\left(
\begin{matrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o}
\end{matrix}
\right) \sim \mathcal{N}
\left(
\left(
\begin{matrix}
U +\gamma_0 k \\B +\gamma_1 k \\V+\gamma_2 k\\R+\gamma_3 k\\I+\gamma_4 k\\
EW_{Si}\\ EW_{Ca}
\end{matrix}
\right)
,C
\right).
\end{equation*}
The mean of the normal distribution is the intrinsic magnitudes plus an extrinsic magnitude change described by a parameter for each individual supernova
$k$ and global parameters $\gamma_i$.  
To constrain the problem we set  $\langle k \rangle=0$.  We also set  $\gamma_2=1$ meaning that $k$ can be interpreted as $A_V$.

As I perform an Bayesian analysis some mention of priors is required.  A flat prior is used for all parameters except
for the covariance matrix $C_c$, which is constructed from a correlation matrix with  $\nu=2$  LKJ prior and standard
deviations with
 $\mu=0.1$, $\sigma=2.5$ Cauchy (Lorentz) distribution prior restricted to non-negative values.
The nature of the LKJ prior is rather mysterious; one thing to note is that for $\nu=2$ the correlation prior is broad but
approaches zero at the extremum.  The prior disfavors uniform correlations allowing grey offsets to be attributed to the $\Delta$ parameters.

Figure~\ref{sigma:fig} shows the confidence regions for 
 the overall scale off the grey offsets
 ``$\Delta$ scale''
and of the standard deviations of the $C_c$ diagonal.  This parameter subset is constrained and the  ``$\Delta$ scale'' parameter
is not highly correlated with the others.  The average
of the covariance matrix draws (there must be something better to use than this) is 
\begin{equation}
\begin{pmatrix}
0.0050 & 0.0026 & 0.0017 & 0.0015 & 0.0010 \\
0.0026 & 0.0038 & 0.0035 & 0.0029 & 0.0013 \\
0.0017 & 0.0035 & 0.0052 & 0.0045 & 0.0032 \\
0.0015 & 0.0029 & 0.0045 & 0.0046 & 0.0038 \\
0.0010 & 0.0013 & 0.0032 & 0.0038 & 0.0055
 \end{pmatrix}.
 \end{equation}
 In terms of colors $U-V$, $B-V$, $V-R$, and $V-I$ the covariance is
\begin{equation}
\begin{pmatrix}
 0.0068 & 0.0025 & -0.0005 & -0.0012 \\
0.0025 & 0.0020 & -0.0000 & 0.0003 \\
-0.0005 & -0.0000 & 0.0008 & 0.0012 \\
-0.0012 & 0.0003 & 0.0012 & 0.0042
  \end{pmatrix}.
 \end{equation}
These correlations indicate that intrinsic color parameters are available.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{output2/sigma_corner.pdf} 
   \caption{Confidence regions for the parameter ``$\Delta$ scale'' that describes the overall scale off the $\Delta$'s
and of the standard deviations of the diagonal of $C_c$.}
   \label{sigma:fig}
\end{figure}

The ideogram for the grey offsets $\Delta$ for all supernovae is shown in Figure~\ref{Delta_hist:fig}.  The intrinsic dispersion
is given by the standard deviation of $0.10$ mag.  The distribution has a broad tail; a model assumption of a normal intrinsic dispersion
would have been inappropriate.
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{output2/Delta_hist.pdf} 
   \caption{Ideogram for the grey offset $\Delta$ for all supernovae.}
   \label{Delta_hist:fig}
\end{figure}

To illustrate how the linear model reduces the dispersion in observed absolute magnitude, we take the
observed $V_o$, and successive correction of $\gamma_2k$,  $\beta EW_{Si}$, and  $\alpha EW_{Ca}$, 
where the 50\%-ile fit value (of the full model) is used for each parameter.  The histograms over the supernovae for these magnitudes are shown in
Figure~\ref{magresidual:fig}.  The benefit of each  linear correction results in the reduction in dispersion.  The amount
of successive reduction in dispersion is similar for all bands.  This is shown for illustration only: in my attempt to reproduce Gerard's previous
work there were no supernovae reserved for out-of-sample model-comparison tests.  AIC and BIC can be computed.
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=4in]{output2/magresidual_V.pdf} 
   \caption{Distributions of $V$-band corrected magnitudes.}
   \label{magresidual:fig}
\end{figure}

The best-fit distributions for the parameters $\alpha$ and $\beta$, which describe the linear relation between absolute magnitude
and $EW_{Ca}$ and $EW_{Si}$ respectively, are shown in Figure~\ref{alphabeta:fig}.  The 1-$\sigma$ 
credible intervals are $\alpha_0=3.0e-3 \pm 8.6e-4$,
$\alpha_1 = 4.7e-4  \pm 7.2e-4$, $\alpha_2 = 5.3e-4  \pm 5.9e-4$, $\alpha_3 = 7.0e-4  \pm 5.0e-4$,
$\alpha_4 = 2.1e-3  \pm 4.4e-4$,  $\beta_0=0.03  \pm 2.8e-3$,
$\beta_1 =0.03  \pm 2.4e-3$, $\beta_2 = 0.03  \pm 2.0e-3$, $\beta_3 =0.02  \pm 1.8e-3$,
$\beta_4 =0.02  \pm 1.6e-3$.
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.9in]{output2/alpha_corner.pdf} 
   \includegraphics[width=2.9in]{output2/beta_corner.pdf} 
   \caption{Distributions for $\alpha$ and $\beta$.}
   \label{alphabeta:fig}
\end{figure}

The best-fit distribution for the parameters $\gamma$  and a comparison with the predictions for extrinsic
magnitude changes being to CCM dust, are shown in Figure~\ref{gamma:fig}.  The 1-$\sigma$ 
credible intervals are $\gamma_0=1.65 \pm 0.05$,
$\gamma_1 = 1.33  \pm 0.03$, $\gamma_2 = 0.78  \pm 0.1$, $\gamma_3 = 0.57  \pm 0.3$.
The CCM dust predictions are based 
on the central wavelengths of the synthetic broad-band filters, and are expected to be shifted slightly.
By eye the model fit results appear consistent with $R_V=3.1$, although $R_V=2.8$ appears to match better.
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.9in]{output2/gamma_corner.pdf} 
   \includegraphics[width=2.9in]{output2/ccm.pdf} 
   \caption{Best-fit distribution for the parameters $\gamma$ and $\beta$, and a comparison with the predictions for extrinsic
magnitude changes being to CCM dust.}
   \label{gamma:fig}
\end{figure}

\begin{equation*}
\left(
\begin{matrix}
U\\B\\V\\R\\I
\end{matrix}
\right) \sim \mathcal{N}
\left(
\Delta +
\left(
\begin{matrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} +\delta_{00} D +\delta_{10} D^2\\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} +\delta_{01} D  +\delta_{11} D^2\\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} +\delta_{02} D +\delta_{12} D^2\\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} +\delta_{03} D +\delta_{13} D^2\\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}+\delta_{04} D +\delta_{14} D^2
\end{matrix}
\right)
,C_{c}
\right).
\end{equation*}
%\begin{equation*}
%D \sim \mathcal{N}(\bar{D},1)
%\end{equation*}
\begin{equation*}
\left(
\begin{matrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o}
\end{matrix}
\right) \sim \mathcal{N}
\left(
\left(
\begin{matrix}
U +\gamma_0 k \\B +\gamma_1 k \\V+\gamma_2 k\\R+\gamma_3 k\\I+\gamma_4 k\\
EW_{Si}\\ EW_{Ca}
\end{matrix}
\right)
,C
\right).
\end{equation*}
$\langle \Delta \rangle=0$, $\langle k \rangle=0$, $\gamma_2=1$, $\langle D \rangle=0$, $\langle D^2 \rangle=1$.

>>>>

\begin{document}
\section{Data}
The data set consists of synthetic $UBVRI$ peak brightnesses and Si and Ca equivalent widths at peak
of 172 SNf supernovae.  The photometry is normalized to place all supernovae at a common distance, as
inferred by redshift.
The data was provided by Manu Gangler.

\section{Model Fits}
\subsection{One-Color Model}
We hypothesize that Type Ia supernova spectral features are predictive of intrinsic
magnitudes.  The intrinsic magnitudes are not perfectly deterministic, but have some intrinsic dispersion.
An extrinsic physical process common to all supernovae can influence the propagation
of light along the line of sight to incluence apparent magnitudes.

\subsubsection{Setup}
Specifically, we assume a linear relationship between the equivalent widths of the Ca and Si spectral features
$EW_{Ca}$ and $EW_{Si}$, and the intrinsic $UBVRI$ magnitudes
with residual Gaussian dispersion with covariance $C_c$.  A grey magnitude offset $\Delta$ is included for each supernova
to capture grey intrinsic dispersion and distance errors introduced in the normalization of the photometry.
The  extrinsic process is also assumed to be linear.  The observables
$U_o, B_o, V_o, R_o, I_o$, $EW_{Ca,o}$ and $EW_{Si,o}$ have Gaussian uncertainty with covariance $C$.  The model can be expressed
\begin{equation*}
\left(
\begin{matrix}
U\\B\\V\\R\\I
\end{matrix}
\right) \sim \mathcal{N}
\left(
\Delta +
\left(
\begin{matrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} \\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si}  \\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} \\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} \\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si}
\end{matrix}
\right)
,C_{c}
\right)
\end{equation*}
\begin{equation*}
\left(
\begin{matrix}
U_o\\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o}\end{matrix}
\right) \sim \mathcal{N}
\left(
\left(
\begin{matrix}
U +\gamma_0 k \\B +\gamma_1 k \\V+\gamma_2 k\\R+\gamma_3 k\\I+\gamma_4 k\\
EW_{Si}\\ EW_{Ca}
\end{matrix}
\right)
,C
\right)
\end{equation*}
The global parameters $c$, $\alpha$, and $\beta$ are the intercepts and slopes of the bilinear relationships between equivalent widths
and intrinsic magnitudes.  The extrinsic piece is split into the properties $\gamma$ of the physical process and the per-supernova 
parameters $k$ that indicate how much of that process each object undergoes.  To break the degeneracies inherent in the model we impose
\begin{equation*}
\langle \Delta \rangle=0, \langle k \rangle=0, \langle k^2 \rangle=1, \gamma_0 \ge 0.
\end{equation*}

Having the intrinsic dispersion $C_c$ as fit parameters seemingly introduces degeneracy in the model, as magnitude and color variation
ascribed to $\Delta$ and $\gamma k$ could also be attributed to intrinsic dispersion.  There are several features of the model
that drives the assignation of variations away from $C_c$:  The probability distribution functions decrease
with increasing $\det{C_c}$ so growing $C_c$ is disfavored; The distributions of $\Delta$ and $k$ are turn out to
be non-Gaussian, and so are not well described by a Normal covariance; The Bayesian prior selected for $C_c$ disfavors
grey offsets.

If interpreted as being due to dust, the extrinsic term could be associated with extinction $A_X = \gamma_X k$.  In
this parlance $E(B-V) = (\gamma_1-\gamma_2) k$ and $R_X = \frac{\gamma_X}{\gamma_1-\gamma_2}$.
Note that these values of $A_X$, $E(B-V)$, and $R_X$ do not depend on the normalization condition $ \langle k^2 \rangle$, i.e.\
are independent of the transformation $\gamma \rightarrow a\gamma$, $k \rightarrow a^{-1} k$.

As I perform an Bayesian analysis some mention of priors is required.  A flat prior is used for all parameters except
for the covariance matrix $C_c$, which is constructed from a correlation matrix with  $\nu=2$  LKJ prior and standard
deviations with
 $\mu=0$, $\sigma=1$ Cauchy (Lorentz) distribution prior restricted to non-negative values.
The nature of the LKJ prior is rather mysterious; one thing to note is that for $\nu=2$ the correlation prior is broad but
approaches zero at the extremum.

\subsubsection{Results}
The 68\% intervals for the global parameters $\alpha$, $\beta$, $\gamma$, and $\sigma = \sqrt{C_{c,ii}}$  are given in Table~\ref{global:tab}.
The two dimensional contours for the parameters of each filter are shown in Figure~\ref{global:fig}.  The Table does not show
the results for $c$ as their values are not significant for the aims of this analysis, the Figure does show $c$ to indicate the convergence
and correlations of these parameters in the fit.

\begin{table}
\centering
\begin{tabular}{|c|c|c|c|c|c|}
\hline
& $X=0$ &1&2&3&4\\ \hline
$\alpha_X$
&
$0.0031^{0.0008}_{-0.0009}$
&
$0.0005^{0.0007}_{-0.0007}$
&
$0.0006^{0.0006}_{-0.0006}$
&
$0.0007^{0.0005}_{-0.0005}$
&
$0.0021^{0.0004}_{-0.0004}$
\\
$\beta_X$
&
$0.0345^{0.0027}_{-0.0029}$
&
$0.0274^{0.0022}_{-0.0025}$
&
$0.0274^{0.0021}_{-0.0021}$
&
$0.0223^{0.0018}_{-0.0018}$
&
$0.0213^{0.0016}_{-0.0017}$
\\
$\gamma_X$
&
$0.3818^{0.0138}_{-0.0132}$
&
$0.3108^{0.0117}_{-0.0113}$
&
$0.2342^{0.0118}_{-0.0115}$
&
$0.1827^{0.0107}_{-0.0111}$
&
$0.1357^{0.0110}_{-0.0119}$
\\
$\sigma_X$
&
$0.0693^{0.0157}_{-0.0152}$
&
$0.0575^{0.0099}_{-0.0089}$
&
$0.0679^{0.0083}_{-0.0080}$
&
$0.0639^{0.0073}_{-0.0069}$
&
$0.0703^{0.0075}_{-0.0075}$
\\
\hline
\end{tabular}
\caption{68\% Intervals for Global Fit Parameters \label{global:tab}}
\end{table}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.5in]{output7/coeff0.pdf} 
   \includegraphics[width=2.5in]{output7/coeff1.pdf} 
   \includegraphics[width=2.5in]{output7/coeff2.pdf} 
      \includegraphics[width=2.5in]{output7/coeff3.pdf} 
         \includegraphics[width=2.5in]{output7/coeff4.pdf} 
            \caption{Distributions for $c$, $\alpha$, $\beta$, $\gamma$, and $\sigma$ in each of the 5 bands.}
   \label{global:fig}
\end{figure}

We find a small but significant non-zero values for $\alpha_0$ and $\alpha_4$, indicating that $EW_{Ca}$ is an indicator of $U$ and $I$
magnitudes.  All bands show larger and significant non-zero values for $\beta$.  This confirms the part of our hypothesis that spectral indicators
are tracers of supernova absolute magnitude.

The best-fit distribution for the parameters $\gamma$  and our measurement of
$\frac{\gamma_X}{\gamma_1-\gamma_2}$  compared with the predictions of CCM dust are shown in Figure~\ref{gamma:fig}:
our fit values for $\gamma$ are consistent with the  $R_V=3.1$ CCM model.  The uncertainties in the 5 measurements
of $\frac{\gamma_X}{\gamma_1-\gamma_2}$ are correlated: measurements of $\gamma_X/\gamma_2$ has significantly smaller
uncertainties and are consistent with $R_V=3.1$.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.4in]{output7/gamma_corner.pdf} 
   \includegraphics[width=3.3in]{output7/ccm.pdf} 
   \caption{Left: Best-fit distribution for the parameters $\gamma$.  Right: 68\% measurements of $\frac{\gamma_X}{\gamma_1-\gamma_2}$ and CCM
   predictions for different values of $R_V$.}
   \label{gamma:fig}
\end{figure}

Non-trivial residual magnitude dispersions are captured in $C_c$.   Figure~\ref{sigma:fig} shows the confidence regions for the
square root of the diagonal elements of $C_c$.
A representative example is given by the matrix comprised by the average of each element in the Monte Carlo chain.
For $UBVRI$ the matrix is
\begin{equation}
\begin{pmatrix}
0.0050 & 0.0026 & 0.0017 & 0.0014 & 0.0009 \\
0.0026 & 0.0034 & 0.0032 & 0.0026 & 0.0011 \\
0.0017 & 0.0032 & 0.0047 & 0.0041 & 0.0029 \\
0.0014 & 0.0026 & 0.0041 & 0.0042 & 0.0035 \\
0.0009 & 0.0011 & 0.0029 & 0.0035 & 0.0050
 \end{pmatrix}.
 \end{equation}
 In terms of colors $U-V$, $B-V$, $V-R$, and $V-I$ the covariance is
\begin{equation}
\begin{pmatrix}
0.0064 & 0.0025 & -0.0004 & -0.0010 \\
0.0025 & 0.0018 & -0.0000 & 0.0003 \\
-0.0004 & -0.0000 & 0.0007 & 0.0011 \\
-0.0010 & 0.0003 & 0.0011 & 0.0039
  \end{pmatrix}.
 \end{equation}
 The $B-V$ and $V-R$ colors have small standard deviations of  0.04 mag and 0.03 mag with no correlation.  The
 residuals not accounted by the
 model are stronger in the $U-V$ and $V-I$ bands.

 \begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.4in]{output7/sigma_corner.pdf} 
   \caption{Best-fit distribution for the parameters $\sigma$, the square root of the diagonal elements of $C_c$.}
   \label{sigma:fig}
\end{figure}

The ideogram for the grey offsets $\Delta$ for all supernovae is shown in Figure~\ref{hist:fig}.  The intrinsic dispersion
is given by the standard deviation of $0.10$ mag.  The distribution is non-Gaussian, with a broad tail. 
\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output7/Delta_hist.pdf} 
   \includegraphics[width=2.8in]{output7/k_hist.pdf} 
   \caption{Ideograms for Left: the grey offset $\Delta$; Right: the color term $k$ for all supernovae.}
   \label{hist:fig}
\end{figure}

The ideogram for the color term  $k$ for all supernovae is shown in Figure~\ref{hist:fig}.The distribution is non-Gaussian, with a broad tail.

To illustrate how the linear model reduces the dispersion in observed absolute magnitude, plot the observed colors as a
function of the observed  $EW_{Ca, o}$, $EW_{Si,o}$ and $B_o-V_o$ in Fig.~\ref{magresidual:fig}.  Overplotted are lines with the
best-fit slope.  The data in the color magnitude diagram indicate that the slope at bluer colors is different from that at redder colors
and that our one-color linear model could be improved upon.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=2.8in]{output7/speccamag.pdf} 
      \includegraphics[width=2.8in]{output7/specsimag.pdf} 
   \includegraphics[width=2.8in]{output7/colormag.pdf} 
   \caption{Observed $UBVRI$ as a function of observed   $EW_{Ca, o}$ (top left), $EW_{Si,o}$ (top right) and $B_o-V_o$ (bottom).
   Overplotted are lines with the best-fit slope from the model fit.}
   \label{magresidual:fig}
\end{figure}

\subsubsection{Notes}
Some equivalent widths go positive.  Is this an emission?

Add Si line velocity.

\subsection{Two-Color Model}
Motivated by the indication that the one-parameter linear model does not fully describe the color-magnitude data, we develop a new model.
Considering that dust does predict a linear extinction in magnitudes, we preserve $k$ as a color parameter and the $\gamma k$ relation
with magnitude.   A second color parameter $D$ is introduced to be quadratically related with magnitude.  Practically I have found
that the data are not sufficient to constrain both the linear and quadratic terms of $D$, so only the latter is considered.

\subsubsection{Setup}
\begin{equation*}
\left(
\begin{matrix}
U\\B\\V\\R\\I
\end{matrix}
\right) \sim \mathcal{N}
\left(
\Delta +
\left(
\begin{matrix}
c_0+\alpha_0 EW_{Ca} + \beta_0 EW_{Si} +\delta_{10} D^2\\
c_1+\alpha_1 EW_{Ca} + \beta_1 EW_{Si} +\delta_{11} D^2\\
c_2+\alpha_2 EW_{Ca} + \beta_2 EW_{Si} +\delta_{12} D^2\\
c_3+\alpha_3 EW_{Ca} + \beta_3 EW_{Si} +\delta_{13} D^2\\
c_4+\alpha_4 EW_{Ca} + \beta_4 EW_{Si} +\delta_{14} D^2
\end{matrix}
\right)
,C_{c}
\right).
\end{equation*}
%\begin{equation*}
%D \sim \mathcal{N}(\bar{D},1)
%\end{equation*}
\begin{equation*}
\left(
\begin{matrix}
U_o  \\B_o\\ V_o\\R_o\\I_o\\EW_{Si, o}\\ EW_{Ca, o}
\end{matrix}
\right) \sim \mathcal{N}
\left(
\left(
\begin{matrix}
U +\gamma_0 k \\B +\gamma_1 k \\V+\gamma_2 k\\R+\gamma_3 k\\I+\gamma_4 k\\
EW_{Si}\\ EW_{Ca}
\end{matrix}
\right)
,C
\right).
\end{equation*}
\begin{equation*}
\langle \Delta \rangle=0, \langle k \rangle=0, \langle k^2 \rangle=1, \langle (D-\bar{D})^2 \rangle=1, \gamma_0 \ge 0.
\end{equation*}
